Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-01 notebooks/expected_and_actual_counts_ratio.Rmd
Add alternative code that can only handle 2 clades and round before calculating p-values

library(data.table)
library(ggplot2)
library(stringr)
library(gtools)
```

@@

# Create a list of the clades to include in comparisons.
clades <- c("20A","20B","20C")

@@

# Add column with the mutation type (clade founder.mutation).
filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
```

@@

@@

```{r head_data}
head(filtered_SARS_mutation, n=5)
```

@@

```{r describe_data}
# How many rows? How many columns?
dim(filtered_SARS_mutation)

# How many phylogenetic clades?
length(unique(filtered_SARS_mutation$clade))

# How many nucleotide sites?
length(unique(filtered_SARS_mutation$nt_site))
```

@@

```{r stacked_bar, echo=FALSE}
# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  geom_bar(position="stack", stat="identity")+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r nucleotide_site, echo=FALSE}
ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
  geom_point()+
  theme_bw()+
  labs(x="", y="")
```

@@

```{r filter_contingency_data}
# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$expected_count, observed_expected=clade_SARS_mutation$observed_expected)
```

@@

For each mutation type (e.g., A -> T) at each nucleotide site, this notebook will generate contingency table analyses for each comparison in the set of clade comparisons. To expand the clade comparison matrix to include mutation types at nucleotide sites, generate a unique list of each mutation type (i.e., three mutations for one clade founder) at each nucleotide site.

```{r unique_mutations}
# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
nt_mutation_list <- unique(paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, sep="_"))
```

@@

```{r combined_clade_comparisons}
# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind", replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))

@@

```{r filter_for_comparisons}
# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
# DOUBLE CHECK THIS: NOT CONVINCED IT IS WORKING
filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'

@@

```{r p_value_correction}
# Identify clades from the two columns in the comparison matrix and combine into one column.
comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,3], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,3], sep="."))

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
comparison_p_values$fold_change <- foldchange(comparison_p_values$clade1_obs_exp, comparison_p_values$clade2_obs_exp)
```

@@

```{r volcano_plot}
# Only plot data for the two clades being compared.
plot_data <- comparison_p_values[which(comparison_p_values$clades=="20B.20C"),]
ggplot(plot_data, aes(x=fold_change, y=-log10(p_value)))+
  geom_point()
```

@@

>>>>

library(data.table)
library(ggplot2)
library(stringr)
```

@@

# Create a list of the clades to include in comparisons.
clades <- c("22A","22B")

@@

# Add column with the mutation type (clade founder.mutation).
#filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
```

@@

```{r round_expected}
# Round expected counts because contingency table analyses (i.e., Fisher's Exact test) can only analyze integers
filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
```

@@

```{r head_data}
#head(filtered_SARS_mutation, n=5)
```

@@

```{r describe_data}
# How many rows? How many columns?
#dim(filtered_SARS_mutation)

# How many phylogenetic clades?
#length(unique(filtered_SARS_mutation$clade))

# How many nucleotide sites?
#length(unique(filtered_SARS_mutation$nt_site))
```

@@

```{r stacked_bar, echo=FALSE}
# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
#ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  #geom_bar(position="stack", stat="identity")+
  #theme_bw()+
  #labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r nucleotide_site, echo=FALSE}
#ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
  #geom_point()+
  #theme_bw()+
  #labs(x="", y="")
```

@@

```{r filter_contingency_data}
# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_mutation, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$rounded_expected, observed_expected=clade_SARS_mutation$observed_expected)
```

@@

@@

```{r combined_clade_comparisons}
# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_SARS_mutation$nt_mutation), each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind", replicate(length(unique(filtered_SARS_mutation$nt_mutation)), clade_comparisons, simplify=FALSE))

@@

```{r filter_for_comparisons}
# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
filtered_comparisons <- comparisons[(comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade),]
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))

@@

```{r p_value_correction}
# Identify clades from the two columns in the comparison matrix and combine into one column.
comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,2], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,2], sep="."))

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
# Double check calculation with Jesse's paper
comparison_p_values$fold_change <- log2(ifelse(comparison_p_values$clade1_obs_exp >= comparison_p_values$clade2_obs_exp, comparison_p_values$clade1_obs_exp/comparison_p_values$clade2_obs_exp, -comparison_p_values$clade2_obs_exp/comparison_p_values$clade1_obs_exp))
```

@@

```{r volcano_plot}
# Only plot data for the two clades being compared.
#plot_data <- comparison_p_values[which(comparison_p_values$clades=="20A.22A"),]
ggplot(comparison_p_values, aes(x=other_fold, y=-log10(p_values_corrected)))+
  geom_point()
```

@@

## Shorted Script That Can Handle 2 Clades
```{r two_clades}
# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds
{start_two <- Sys.time()
SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
country <- "all"
clades <- c("20A","20B")
min_expected <- 0
pseudo <- 0.5
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
clade1 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[1]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
clade2 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[2]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
m <- merge(clade1, clade2, by="nt_mutation")
filtered_comparisons <- data.frame(clade1=paste(m$nt_mutation, m$clade.x, sep="_"), clade2=paste(m$nt_mutation, m$clade.y, sep="_"), clade1_observed=m$actual_count.x, clade1_expected=m$expected_count.x, clade2_observed=m$actual_count.y, clade2_expected=m$expected_count.y, clade1_obs_exp=m$observed_expected.x, clade2_obs_exp=m$observed_expected.y)
end_two <- Sys.time()}
end_two - start_two
two <- filtered_comparisons
```
